Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-18 checking-executions.tex
sequential executions
example for the sequential procedure

Checking whether a data-differentiated sequential execution belongs to $\seqPQ$ could be implemented by checking membership into the language accepted by $LTS??$. The recursive procedure $\textit{Check-PQ}$ outlined in Algorithm~\ref{alg:seq_check} is an alternative to this membership test. Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations. Since the $\textit{rm}(\textit{empty})$ operations are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(a,p)$ operation before $o$ is matched to a $\textit{rm}(a)$ operation which also occurs before $o$. Once there are no more $\textit{rm}(\textit{empty})$ operations, the procedure selects either a $\textit{put}$ operation adding a value with maximal priority which is not removed, or a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. Their correctness can be checked using a similar argument as for $\textit{rm}(\textit{empty})$ operations.

@@

\vspace{-3mm}
\noindent
where $p\prec \textit{priorities}(e)$ when $p\prec p'$ for some $p'\in \textit{priorities}(e)$ (and similarly for $p\prec \textit{unmatched-priorities}(e)$ or $p\preceq \textit{unmatched-priorities}(e)$), 
$\textit{matched}_\prec(e,p)$ holds when each value with priority strictly smaller than $p$ is removed in $e$, and $\textit{matched}(e)$ holds when $\textit{matched}_\prec(e,p)$ holds for each $p\in\mathbb{P}$.

@@

TODO MODIFY THE FOLLOWING EXAMPLE

Let us use an example to show how $\textit{check-PQ}$ works. Here we use $w_1 \xrightarrow{R} w_2$ to denote that in $\textit{check-PQ}(w_1)$, we choose the branch of $R$ and recursively call $\textit{check-PQ}(w_2)$.

\begin{example}\label{example:generate extended priority queue executions}
Given priorities $p_1,p_2,p_3$ with orders $p_1 \prec p_2$ and $p_1 \prec p_3$,

\noindent $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)$

$\xrightarrow{\textit{EPQ}_3}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)$

$\xrightarrow{\textit{EPQ}_1}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(b,p_1)$

$\xrightarrow{\textit{EPQ}_2}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(b,p_1)$

$\xrightarrow{\textit{EPQ}_1}$ $\textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{put}(b,p_1)$

$\xrightarrow{\textit{EPQ}_2}$ $\textit{put}(a,p_1) \cdot \textit{rm}(a)$

$\xrightarrow{\textit{EPQ}_1}$ $\epsilon$
\end{example}

@@

\begin{lemma}
Algorithm~\ref{alg:seq_check} returns $\mathsf{true}$ on an execution $e$ iff $e\in \seqPQ$.
\end{lemma}

>>>>

Checking whether a data-differentiated sequential execution belongs to $\seqPQ$ could be implemented by checking membership into the language accepted by $LTS??$. The recursive procedure $\textit{Check-PQ}$ outlined in Algorithm~\ref{alg:seq_check} is an alternative to this membership test. Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations.

We explain how the procedure works on the following execution:
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Therefore, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of 
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
When there are no more $\textit{rm}(\textit{empty})$ operations, the procedure selects either a $\textit{put}$ operation adding a value with maximal priority which is not removed, or a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For instance, since $p_2$ is a maximal priority, it may select the operation $\textit{put}(d,p_2)$. This operation is trivially correct 
and the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of 
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
Now, all the maximal priority values are removed and the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations, for instance, $\textit{put}(c,p_2)$ and $\textit{rm}(c)$. In this case, the value returned by $\textit{rm}$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, the only value of priority smaller than $p_2$ is $a$ and satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

@@

\vspace{-3mm}
\noindent
where $p\prec \textit{priorities}(e)$ when $p\prec p'$ for some $p'\in \textit{priorities}(e)$ (and similarly for $p\prec \textit{unmatched-priorities}(e)$ or $p\preceq \textit{unmatched-priorities}(e)$), 
$\textit{matched}_\prec(e,p)$ holds when each value with priority strictly smaller than $p$ is removed in $e$, and $\textit{matched}(e)$ holds when $\textit{matched}_\prec(e,p)$ holds for each $p\in\mathbb{P}$. Compared to the example presented at the beginning of the section, these predicates take into consideration that multiple values with the same priority are removed in FIFO order: the predicate $\mathsf{MatchedMaxPrioritySeq}(e,x)$ holds when $x$ is the last value with priority $p$ added in $e$.

@@

%TODO MODIFY THE FOLLOWING EXAMPLE
%
%Let us use an example to show how $\textit{check-PQ}$ works. Here we use $w_1 \xrightarrow{R} w_2$ to denote that in $\textit{check-PQ}(w_1)$, we choose the branch of $R$ and recursively call $\textit{check-PQ}(w_2)$.
%
%\begin{example}\label{example:generate extended priority queue executions}
%Given priorities $p_1,p_2,p_3$ with orders $p_1 \prec p_2$ and $p_1 \prec p_3$,
%
%\noindent $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_3}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_1}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_2}$ $\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_1}$ $\textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{put}(b,p_1)$
%
%$\xrightarrow{\textit{EPQ}_2}$ $\textit{put}(a,p_1) \cdot \textit{rm}(a)$
%
%$\xrightarrow{\textit{EPQ}_1}$ $\epsilon$
%\end{example}

@@

\begin{restatable}{lemma}{EPQRulesAndSemantics}
\label{lemma:EPQ rules and semantics}
Algorithm~\ref{alg:seq_check} returns $\mathsf{true}$ on an execution $e$ iff $e\in \seqPQ$.
\end{restatable}

%
%\begin{lemma}
%Algorithm~\ref{alg:seq_check} returns $\mathsf{true}$ on an execution $e$ iff $e\in \seqPQ$.
%\end{lemma}
